---
title: "NicheNet example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the
**NicheNet** tool and have a look at the output it produces. More information
about **NicheNet** can be found at https://github.com/saeyslab/nichenetr.

```{r libraries}
library("nichenetr")
```

Input
=====

**NicheNet** takes four inputs, a matrix of ligand targets, a ligand-receptor
network, an scRNA-seq dataset and a gene set of interest.

Ligand-target matrix
--------------------

This matrix contains the prior potential that a particular ligand might regulate
the expression of a specific target gene. Here is a snippet of the matrix:

```{r input-ligand-target}
# From https://zenodo.org/record/3260758/files/ligand_target_matrix.rds
ligand_targets <- read_rds(
    fs::path(PATHS$NicheNet_in, "ligand_target_matrix.Rds")
)

pander(ligand_targets[1:5, 1:5])
```

The full matrix has **`r nrow(ligand_targets)`** rows (targets) and
**`r ncol(ligand_targets)`** columns (ligands).

Ligand-receptor network
-----------------------

This is a database of ligand-receptor pairs with information about their source.

```{r input-network}
# From https://zenodo.org/record/3260758/files/lr_network.rds
lr_network <- read_rds(fs::path(PATHS$NicheNet_in, "lr_network.Rds"))

skim(lr_network)
```

Expression data
---------------

The example dataset is from cancer-associated fibroblasts (CAFs) in the head and
neck squamous cell carcinoma (HNSCC) tumour microenvironment.

```{r input-expression}
# From https://zenodo.org/record/3260758/files/hnscc_expression.rds
hnscc <- read_rds(fs::path(PATHS$NicheNet_in, "hnscc_expression.Rds"))
```

This dataset is provided as a list with the following items:
`r paste("**", names(hnscc), "**", sep = "", collapse = ", ")`.

### Expression matrix

The first item is the expression matrix.

```{r input-expression-matrix}
exprs_mat <- hnscc$expression

pander(exprs_mat[1:5, 1:5])
```

The expression matrix has **`r nrow(exprs_mat)`** rows (cells) and
**`r ncol(exprs_mat)`** columns (genes).

### Sample information

There is also some metadata about the cells.

```{r input-expression-samples}
sample_info <- hnscc$sample_info

skim(sample_info)
```

### Expressed genes

The final item is a vector of the names of expressed genes.

```{r input-expression-genes}
skim(hnscc$expressed_genes)
```

Gene set
--------

Because we are looking at how CAFs influence cancer growth we will use a
signature for p-EMT.

```{r input-geneset}
# From https://zenodo.org/record/3260758/files/pemt_signature.txt
geneset <- read_tsv(
    fs::path(PATHS$NicheNet_in, "pemt_signature.txt"),
    col_types = cols(
        gene = col_character()
    ),
    col_names = "gene"
) %>%
    pull(gene) %>%
    .[. %in% rownames(ligand_targets)]

skim(geneset)
```

Define expressed genes
======================

The first step in the analysis is to define which genes are expressed in the
sender and receiver cell populations. In this example the CAFs are defined to be
the senders and tumour cells (from high quality tumours) are defined to be
receivers.

We also need to set a threshold for deciding a gene is "expressed". Here we use
the following formula:

$$Ea_i = log_2((\frac{1}{k}\sum_{i = 1}^{k} TPM_i) + 1) >= 4$$

> **NOTE:** For UMI data (10x Chromium) the authors don't use this formula and
> instead suggest a threshold of non-zero expression in at least 10% of cells.

```{r expressed genes}
# Low-quality tumours to remove
tumours_remove <- c("HN10", "HN", "HN12", "HN13", "HN24", "HN7", "HN8", "HN23")

CAF_ids <- sample_info %>%
    filter(
        `Lymph node` == 0 &
            !(tumor %in% tumours_remove) &
            `non-cancer cell type` == "CAF"
    ) %>%
    pull(cell)

malignant_ids <- sample_info %>%
    filter(
        `Lymph node` == 0 &
            !(tumor %in% tumours_remove) &
            `classified  as cancer cell` == 1
    ) %>%
    pull(cell)

expressed_sender <- exprs_mat[CAF_ids, ] %>%
    apply(2, function(x) {10 * (2 ** x - 1)}) %>%
    apply(2, function(x) {log2(mean(x) + 1)}) %>%
    .[. >= 4] %>%
    names()
expressed_receiver <- exprs_mat[malignant_ids, ] %>%
    apply(2, function(x) {10 * (2 ** x - 1)}) %>% 
    apply(2, function(x) {log2(mean(x) + 1)}) %>%
    .[. >= 4] %>%
    names()
```

After this quality control we have selected **`r length(CAF_ids)`** CAFs
(senders) and **`r length(malignant_ids)`** tumour cells (receivers). There are
**`r length(expressed_sender)`** genes expressed in the sender cells and
**`r length(expressed_receiver)`** genes expressed in the receiver cells.

Background genes
================

We already have the gene set of interest that we want to look at but we also
need to define a background set of genes. For this we use all genes expressed in
the malignant (receiver) cells which are also in the ligand-target matrix.

```{r background}
background_genes <- expressed_receiver %>%
    .[. %in% rownames(ligand_targets)]
```

This gives us a set of **`r length(background_genes)`** background genes.

Potential ligands
=================

We determine a set of potential ligands by selection those that are expressed
by CAFs (sender) and bind a receptor expressed in malignant cells (receiver).

```{r ligands}
ligands <- lr_network %>%
    pull(from) %>%
    unique()

expressed_ligands <- intersect(ligands, expressed_sender)

receptors <- lr_network %>%
    pull(to) %>%
    unique()

expressed_receptors <- intersect(receptors, expressed_receiver)

lr_network_expressed <- lr_network %>%
    filter(
        from %in% expressed_ligands &
            to %in% expressed_receptors
    )

potential_ligands <- lr_network_expressed %>%
    pull(from) %>%
    unique()
```

The filtered database contains **`r nrow(lr_network_expressed)`** expressed
ligand-receptor pairs with **`r length(potential_ligands)`** potential ligands.

Ligand activity analysis
========================

Now we have all the input data sorted we can run **NicheNet**. The first
analysis assesses ligand activity by calculating how well each CAF-ligand can
predict that a gene belongs to the p-EMT gene set compared to the background
genes.

```{r activity}
ligand_activities <- predict_ligand_activities(
    geneset                    = geneset, 
    background_expressed_genes = background_genes,
    ligand_target_matrix       = ligand_targets,
    potential_ligands          = potential_ligands
)

skim(ligand_activities)
```

There are various scores given as results here but the authors suggest using
the Pearson correlation to select ligands. We select 20 ligands with highest
correlation.

```{r best-activity}
best_ligands <- ligand_activities %>%
    top_n(20, pearson) %>%
    arrange(-pearson) %>%
    pull(test_ligand)
```

These ligands are:
`r glue_collapse(glue("**{best_ligands}**"), sep = ", ", last = " and ")`

The choice of to use 20 ligands is somewhat arbitary and is likely to be
different for different settings. For a real analysis the authors suggest
choosing a threshold by looking at the distribution of correlations.

```{r activity-plot}
pearson_thresh <- ligand_activities %>%
    top_n(20, pearson) %>%
    pull(pearson) %>%
    min()

ggplot(ligand_activities, aes(x = pearson)) + 
    geom_histogram(color = "black", fill = "darkorange")  +
    geom_vline(
        xintercept = pearson_thresh,
        color      = "red",
        linetype   = "dashed",
        size       = 1
    ) + 
    labs(
        x = "ligand activity (PCC)",
        y = "# ligands"
    ) +
    theme_classic()
```

Infer target genes
==================

Once we have a set of active ligands we can look at the regulatory potential
between ligands and downstream targets. We only look at interactions between the
top 20 ligands and genes that belong to the gene set and are in the top 250
most strongly predicted targets of one of the selected ligands. The targets are
selected based on predictions in the general prior model so are not specific to
this dataset. Genes that are not a top target of one of the selected ligands
will not be shown.

```{r active-links}
active_links <- best_ligands %>%
    lapply(
        get_weighted_ligand_target_links,
        geneset              = geneset,
        ligand_target_matrix = ligand_targets,
        n                    = 250
    ) %>%
    bind_rows()

skim(active_links)
```

For visualisation potential scores are set to zero if they were below the 0.25
quantile for top targets of that ligand in the ligand-target matrix.

```{r target-heatmap}
links_mat = prepare_ligand_target_visualization(
    ligand_target_df     = active_links,
    ligand_target_matrix = ligand_targets,
    cutoff               = 0.25
)

order_ligands <- intersect(best_ligands, colnames(links_mat)) %>% rev()
order_targets <- active_links$target %>% unique()

vis_ligand_target <- links_mat[order_targets, order_ligands] %>% t()

vis_ligand_target %>%
    make_heatmap_ggplot(
        "Prioritized CAF-ligands",
        "p-EMT genes in malignant cells",
        color           = "purple",
        legend_position = "top",
        x_axis_position = "top",
        legend_title    = "Regulatory potential"
    ) +
    scale_fill_gradient2(
        low = "whitesmoke",
        high = "purple",
        breaks = c(0,0.005,0.01)
    ) +
    theme(axis.text.x = element_text(face = "italic"))
```

The cutoffs for visualisation are arbitary and the authors suggest testing
several cutoffs. Considering more targets would identify more interactions but
with less confidence. Lowering the quantil cutoff would result in a denser
heatmap.

Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information {.unnumbered}
-------------------
