---
title: "COMUNET example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the
**COMUNET** package and have a look at the output it produces. More information
about **COMUNET** can be found at https://github.com/ScialdoneLab/COMUNET.

```{r libraries}
library("COMUNET")
```

Input
=====

The **COMUNET** package performs downstream analysis based on the results of 
any algorithm which produces a matrix of weights representing the strength of
interactions between two cell type from a ligand-receptor pair.

For their tutorials the authors have used the output produced by
**CellPhoneDB**.

CellPhoneDB database
--------------------

We require two of the files which make up the **CellPhoneDB** database.

### complex_input.csv

This file contains information about the complexes in the **CellPhoneDB**
database.

```{r input-complex}
complex_input <- read_csv(
    fs::path(
        PATHS$cellphonedb_in,
        "database_v2.0.0",
        "data",
        "complex_input.csv"
    ),
    col_types = cols(
        complex_name       = col_character(),
        uniprot_1          = col_character(),
        uniprot_2          = col_character(),
        uniprot_3          = col_character(),
        uniprot_4          = col_logical(),
        transmembrane      = col_logical(),
        peripheral         = col_logical(),
        secreted           = col_logical(),
        secreted_desc      = col_character(),
        secreted_highlight = col_logical(),
        receptor           = col_logical(),
        receptor_desc      = col_character(),
        integrin           = col_logical(),
        other              = col_logical(),
        other_desc         = col_character(),
        pdb_id             = col_character(),
        pdb_structure      = col_character(),
        stoichiometry      = col_character(),
        comments_complex   = col_character()
    )
) %>%
    mutate(complex_name = gsub("_" , " " , complex_name))

skim(complex_input)
```

### gene_input.csv

This file contains information about the genes in the **CellPhoneDB** database.

```{r input-gene}
gene_input <- read_csv(
    fs::path(
        PATHS$cellphonedb_in,
        "database_v2.0.0",
        "data",
        "gene_input.csv"
    ),
    col_types = cols(
        gene_name   = col_character(),
        uniprot     = col_character(),
        hgnc_symbol = col_character(),
        ensembl     = col_character()
    )
)

skim(gene_input)
```

CellPhoneDB output
------------------

Output files from running **CellPhoneDB**

### significant_means.txt

Information about each ligand-receptor pair as well as scores for each pair of
cell types calculated by **CellPhoneDB**.

```{r input-means}
means <- read_tsv(
    fs::path(PATHS$COMUNET_in, "mouse", "significant_means.txt"),
    col_types = cols(
        .default = col_double(),
        id_cp_interaction   = col_character(),
        interacting_pair    = col_character(),
        partner_a           = col_character(),
        partner_b           = col_character(),
        gene_a              = col_character(),
        gene_b              = col_character(),
        secreted            = col_logical(),
        receptor_a          = col_logical(),
        receptor_b          = col_logical(),
        annotation_strategy = col_character(),
        is_integrin         = col_logical()
    )
)

skim(means)
```

Convert CellPhoneDB output
==========================

The first step is to convert the output from **CellPhoneDB** into the format
used by **COMUNET**. Instead of the single matrix provided by **CellPhoneDB**,
**COMUNET** needs a matrix for each ligand-receptor pair where the rows are
sending cell types and the columns are receiving cell types.

```{r convert}
prepped_means <- means %>%
    as.data.frame() %>%
    distinct(interacting_pair, .keep_all = TRUE)
rownames(prepped_means) <- prepped_means$interacting_pair

interactions <- convert_CellPhoneDB_output(
    CellPhoneDB_output = prepped_means,
    complex_input      = complex_input,
    gene_input         = gene_input
)
```

The result of the conversion function is a list with
**`r length(interactions)`** items.

Weight array
------------

The first item is a three-dimensional array with the weights calculated by
**CellPhoneDB**. There are **`r dim(interactions$weight_array)[3]`** of these
matrices, here is an example of the first one.

```{r convert-weights}
pander(interactions$weight_array[, , 1])
```

Ligand-receptor pairs
---------------------

The second item contains information about ligand-receptor pairs.

```{r convert-pairs}
skim(interactions$ligand_receptor_pair_df)
```

Nodes
-----

The final item is a vector of the names of all the cell types:
`r glue_collapse(glue("**{interactions$nodes}**"), sep = ", ")`

Clustering interaction partners
===============================

The aim of this analysis is to find ligand-receptor pairs that interact in
similar ways. This is done by clustering the communication graphs of the
ligand-receptor pairs.

```{r clustering}
lrp_clusters <- lrp_clustering(
    weight_array            = interactions$weight_array,
    ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
    nodes                   = interactions$nodes
)
```

Output
------

This step produces a list with **`r length(lrp_clusters)`** items.

### Dissimilarity matrix

Matrix with distances between each ligand-receptor pair. Here is a small
example.

```{r clustering-dissim}
pander(lrp_clusters$dissim_matrix[1:5, 1:5])
```

### Clusters

Cluster assignments for each ligand-receptor pair.

```{r clustering-clusters}
kable(head(lrp_clusters$clusters))
```

### Cluster weight array

The average interaction weights between cell types by cluster. There are
**`r dim(lrp_clusters$weight_array_by_cluster)[3]`** of these matrices, here is
an example of the first one. 

```{r clustering-weights}
pander(lrp_clusters$weight_array_by_cluster[, , 1])
```

Visualisation
-------------

We can visualise the results in different ways.

### Heatmap

We can plot a heatmap of the clustered ligand-receptor pairs.

```{r clustering-heatmap}
plot_cluster_heatmap(
    dissim_matrix = lrp_clusters$dissim_matrix,
    lrp_clusters  = lrp_clusters$clusters
)
```

### UMAP

We can also make a UMAP plot showing the pairs in a reduced dimensional space.

```{r clustering-umap}
plot_cluster_UMAP(
    ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
    dissim_matrix           = lrp_clusters$dissim_matrix,
    lrp_clusters            = lrp_clusters$clusters
)
```

### Communication pattern

The average communication between cell types for each cluster can be shown as a
graph. Here are examples for the first three clusters.

```{r clustering-pattern}
for(cluster_idx in c(1:3)){
    cluster <- paste("cluster", cluster_idx)
    plot_communication_graph(
        LRP = cluster,
        weight_array            = lrp_clusters$weight_array_by_cluster[, , cluster],
        ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
        nodes                   = interactions$nodes,
        is_cluster              = TRUE
    )
}
```

### Pairs

We can also look at the specific ligand-receptor pairs in a cluster.

```{r clustering-pairs}
for(cluster_idx in c(1:3)) {
    plot_lig_rec(
        cluster_of_interest     = cluster_idx,
        lrp_clusters            = lrp_clusters$clusters,
        ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
        node_label_cex          = 0.5
    )
}
```

Pattern search
==============

**COMUNET** can also be used to search for specific patterns of interactions.
Here we search for interactions from a specific cell type to all other cell
types.

First we construct a matrix describing the pattern we are interested in.

```{r pattern}
communicating_nodes <- c(
    "exVE_to_EPI", "exVE_to_Mes", "exVE_to_TE", "exVE_to_emVE", "exVE_to_exVE"
)

pattern <- make_pattern_matrix(
    communicating_nodes = communicating_nodes,
    nodes               = interactions$nodes
)

pander(pattern)
```

We can also visualise this pattern to check it is correct.

```{r pattern-vis}
plot_communication_graph(
    LRP = "My pattern of interest",
    weight_array            = pattern,
    ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
    nodes                   = interactions$node,
    is_pattern              = TRUE
)
```

Now we can search for this pattern. The result is a dissimilarity to the search
pattern for each ligand-receptor pair.

```{r pattern-search}
patterns <- pattern_search(
    pattern_adj_matrix      = pattern,
    weight_array            = interactions$weight_array,
    ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
    nodes                   = interactions$nodes
)

skim(patterns)
```

We can visualise examples of some ligand-receptor pairs along with their
dissimilarity to the search pattern.

```{r pattern-pairs}
for (pair in c("IGF2:IGF2R", "EFNB1:EPHA4", "IGF2:IGF1R")) {
    plot_communication_graph(
        LRP                     = pair,
        weight_array            = interactions$weight_array,
        ligand_receptor_pair_df = interactions$ligand_receptor_pair_df,
        nodes                   = interactions$node,
        subtitle                = paste(
            "dissimilarity:", patterns[pair,"dissimilarity"]
        )
    )    
}
```

Comparative analysis
====================

We can also use **COMUNET** to compare the interaction network between two
conditions. For this analysis we use a second dataset that includes AML samples
before and after treatment.

```{r read-AML}
cond1 <- "AML328_d0"
cond2 <- "AML328_d29"

cond1_means <- read_tsv(
    fs::path(PATHS$COMUNET_in, "AML", "means_d0.txt"),
    col_types = cols(
        .default = col_double(),
        id_cp_interaction   = col_character(),
        interacting_pair    = col_character(),
        partner_a           = col_character(),
        partner_b           = col_character(),
        gene_a              = col_character(),
        gene_b              = col_character(),
        secreted            = col_logical(),
        receptor_a          = col_logical(),
        receptor_b          = col_logical(),
        annotation_strategy = col_character(),
        is_integrin         = col_logical()
    )
)

cond2_means <- read_tsv(
    fs::path(PATHS$COMUNET_in, "AML", "means_d29.txt"),
    col_types = cols(
        .default = col_double(),
        id_cp_interaction   = col_character(),
        interacting_pair    = col_character(),
        partner_a           = col_character(),
        partner_b           = col_character(),
        gene_a              = col_character(),
        gene_b              = col_character(),
        secreted            = col_logical(),
        receptor_a          = col_logical(),
        receptor_b          = col_logical(),
        annotation_strategy = col_character(),
        is_integrin         = col_logical()
    )
)

cond1_prepped_means <- cond1_means %>%
    as.data.frame() %>%
    distinct(interacting_pair, .keep_all = TRUE)
rownames(cond1_prepped_means) <- cond1_prepped_means$interacting_pair

cond2_prepped_means <- cond2_means %>%
    as.data.frame() %>%
    distinct(interacting_pair, .keep_all = TRUE)
rownames(cond2_prepped_means) <- cond2_prepped_means$interacting_pair

cond1_interactions <- convert_CellPhoneDB_output(
    CellPhoneDB_output = cond1_prepped_means,
    complex_input      = complex_input,
    gene_input         = gene_input
)

cond2_interactions <- convert_CellPhoneDB_output(
    CellPhoneDB_output = cond2_prepped_means,
    complex_input      = complex_input,
    gene_input         = gene_input
)
```

First we check the overlap in ligand-receptor pairs in the two conditions.

```{r overlap}
cond1_pairs <- cond1_interactions$ligand_receptor_pair_df$pair
cond2_pairs <- cond2_interactions$ligand_receptor_pair_df$pair

inter <- intersect(cond1_pairs, cond2_pairs)
cond1_only <- setdiff(cond1_pairs, cond2_pairs)
cond2_only <- setdiff(cond2_pairs, cond1_pairs)
```

There are **`r length(inter)`** pairs present in both conditions,
**`r length(cond1_only)`**, present only in the first condition and
**`r length(cond2_only)`** present in only the second condition.

Just because the pairs are present doesn't mean they are interacting in the same
way. To find that out we need to run the analysis.

```{r compare}
result <- comparative_analysis(
    cond1_weight_array            = cond1_interactions$weight_array,
    cond2_weight_array            = cond2_interactions$weight_array,
    cond1_ligand_receptor_pair_df = cond1_interactions$ligand_receptor_pair_df,
    cond2_ligand_receptor_pair_df = cond2_interactions$ligand_receptor_pair_df,
    cond1_nodes                   = cond1_interactions$nodes,
    cond2_nodes                   = cond2_interactions$nodes,
    cond1_name                    = cond1,
    cond2_name                    = cond2
)
```

Output
------

The output of the comparison function is a list with **`r length(result)`**
items.

### Pairs

The first item describes the ligand-receptor pairs, which conditions that are
present in and the dissimilarity between the conditions.

```{r comparison-pairs}
skim(result$sorted_LRP_df)
```

### Dissimilarity

The second output is a dissimilarity matrix where the rows are ligand-receptor
pairs in condition 1 and and the columns are ligand-receptor pairs in condition
2.

```{r comparison-dissim}
pander(result$dissim_cond1_cond2[1:5, 1:5])
```

Visualisation
-------------

### Heatmap

We can plot a heatmap of the dissimilarity between conditions.

```{r comparison-heatmap}
plot_dissimilarity_heatmaps(
    dissim_cond1_cond2 = result$dissim_cond1_cond2,
    sorted_LRP_df      = result$sorted_LRP_df,
    cond1_name         = cond1,
    cond2_name         = cond2
)
```

### Graphs

Graphs can be used to show the communication networks for a ligand-receptor
pair. Let's compare the graphs between conditions for a set of example pairs.

#### Most similar

```{r comparison-most}
most_similar <- result$sorted_LRP_df %>%
    filter(
        presence      == "shared",
        dissimilarity == min(dissimilarity)
    )
```

The most similar pair is **`r most_similar$pair`** with a dissimilarity of
**`r most_similar$dissimilarity`**.

```{r comparison-most-graph}
plot_communication_graph(
    LRP                     = most_similar$pair,
    weight_array            = cond1_interactions$weight_array,
    ligand_receptor_pair_df = cond1_interactions$ligand_receptor_pair_df,
    nodes                   = cond1_interactions$node,
    title                   = most_similar$pair,
    subtitle                = cond1
)

plot_communication_graph(
    LRP                     = most_similar$pair,
    weight_array            = cond2_interactions$weight_array,
    ligand_receptor_pair_df = cond2_interactions$ligand_receptor_pair_df,
    nodes                   = cond2_interactions$node,
    title                   = most_similar$pair,
    subtitle                = cond2
)
```

#### Least similar

```{r comparison-least}
least_similar <- result$sorted_LRP_df %>%
    filter(
        presence      == "shared",
        dissimilarity == max(dissimilarity)
    ) %>%
    top_n(1, pair)
```

The least similar pair is **`r least_similar$pair`** with a dissimilarity of
**`r least_similar$dissimilarity`**.

```{r comparison-least-graph}
plot_communication_graph(
    LRP                     = least_similar$pair,
    weight_array            = cond1_interactions$weight_array,
    ligand_receptor_pair_df = cond1_interactions$ligand_receptor_pair_df,
    nodes                   = cond1_interactions$node,
    title                   = least_similar$pair,
    subtitle                = cond1
)

plot_communication_graph(
    LRP                     = least_similar$pair,
    weight_array            = cond2_interactions$weight_array,
    ligand_receptor_pair_df = cond2_interactions$ligand_receptor_pair_df,
    nodes                   = cond2_interactions$node,
    title                   = least_similar$pair,
    subtitle                = cond2
)
```

#### Condition 1 only

An example of a pair only in condition 1 is **`r cond1_only[1]`**.

```{r comparison-cond1-graph}
plot_communication_graph(
    LRP                     = cond1_only[1],
    weight_array            = cond1_interactions$weight_array,
    ligand_receptor_pair_df = cond1_interactions$ligand_receptor_pair_df,
    nodes                   = cond1_interactions$node,
    title                   = cond1_only[1],
    subtitle                = cond1
)

plot_communication_graph(
    LRP                     = cond1_only[1],
    weight_array            = cond2_interactions$weight_array,
    ligand_receptor_pair_df = cond2_interactions$ligand_receptor_pair_df,
    nodes                   = cond2_interactions$node,
    title                   = cond1_only[1],
    subtitle                = cond2
)
```

#### Condition 2 only

An example of a pair only in condition 2 is **`r cond2_only[1]`**.

```{r comparison-cond2-graph}
plot_communication_graph(
    LRP                     = cond2_only[1],
    weight_array            = cond1_interactions$weight_array,
    ligand_receptor_pair_df = cond1_interactions$ligand_receptor_pair_df,
    nodes                   = cond1_interactions$node,
    title                   = cond2_only[1],
    subtitle                = cond1
)

plot_communication_graph(
    LRP                     = cond2_only[1],
    weight_array            = cond2_interactions$weight_array,
    ligand_receptor_pair_df = cond2_interactions$ligand_receptor_pair_df,
    nodes                   = cond2_interactions$node,
    title                   = cond2_only[1],
    subtitle                = cond2
)
```

Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
lrp_clusters$dissim_matrix %>%
    as.data.frame() %>%
    rownames_to_column("Pair") %>%
    write_tsv(fs::path(OUT_DIR, "cluster_dissimilarity.tsv"))

tibble(Pair = names(lrp_clusters$clusters), Cluster = lrp_clusters$clusters) %>%
    write_tsv(fs::path(OUT_DIR, "clusters.tsv"))

write_rds(
    lrp_clusters$weight_array_by_cluster,
    fs::path(OUT_DIR, "cluster_weights.Rds")
)

write_tsv(patterns, fs::path(OUT_DIR, "patterns.tsv"))

write_tsv(result$sorted_LRP_df, fs::path(OUT_DIR, "comparison_pairs.tsv"))

result$dissim_cond1_cond2 %>%
    as.data.frame() %>%
    rownames_to_column("Condition1") %>%
    write_tsv(fs::path(OUT_DIR, "comparison_dissimilarity.tsv"))

kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR),
        download_link("cluster_dissimilarity.tsv", OUT_DIR),
        download_link("clusters.tsv", OUT_DIR),
        download_link("cluster_weights.Rds", OUT_DIR),
        download_link("patterns.tsv", OUT_DIR),
        download_link("comparison_pairs.tsv", OUT_DIR),
        download_link("comparison_dissimilarity.tsv", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis",
        "Cluster dissimilarity matrix",
        "Cluster assignments for pairs",
        "Cluster average weights array",
        "Pattern dissimilarity for pairs",
        "Comparison information about pairs",
        "Comparison dissimilarity matrix"
    )
))
```

Session information {.unnumbered}
-------------------
