---
title: "CellChat example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the
**CellChat** package and have a look at the output it produces. More information
about **CellChat** can be found at https://github.com/sqjin/CellChat.

```{r libraries}
library("CellChat")
```

Input
=====

Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information {.unnumbered}
-------------------
