---
title: "CellChat example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the
**CellChat** package and have a look at the output it produces. More information
about **CellChat** can be found at https://github.com/sqjin/CellChat.

```{r libraries}
library("CellChat")
library("ggalluvial")
```

Input
=====

**CellChat** requires two user inputs: an expression matrix and either cell-type
labels or a low-dimensional embedding. Here we use the label-based mode. The
other inputs to **CellChat** are the ligand-receptor database and
protein-protein interaction network included in the package.

```{r load}
load(fs::path(PATHS$CellChat_in, "data_wound_CellChat.rda"))
```

Expression matrix
-----------------

The expression matrix should be normalised with genes as rows and cells as
columns. Here is a snippet of the example dataset:

```{r input-expr}
expr <- data_wound$data

pander(as.matrix(expr[1:5, 1:5]))
```

The full matrix has **`r nrow(expr)`** rows (genes) and **`r ncol(expr)`**
columns (cells).

Labels
------

The second input is a `data.frame` with cell type labels for each cell.

```{r input-labels}
labels <- data.frame(
    group     = data_wound$labels,
    row.names = names(data_wound$labels)
)
group_size <- as.numeric(table(labels$group))

skim(labels)
```

Database
--------

The database is available as a list. The tutorial suggests using only the
interactions annotated with "Secreted Signalling".

```{r input-database}
CellChatDB <- CellChatDB.mouse
CellChatDB <- subsetDB(CellChatDB, search = "Secreted Signaling")
```

The database list contains **`r length(CellChatDB)`** items: 
`r glue_collapse(glue("**{names(CellChatDB)}**"), sep = ", ", last = " and ")`

### Interaction

The first item in the list is a `data.frame` with information about
interactions.

```{r input-database-interaction}
skim(CellChatDB$interaction)
```

### Complex

The second item contain information about complexes.

```{r input-database-complex}
skim(CellChatDB$complex)
```

### Cofactor

The next item contains information about cofactors.

```{r input-database-cofactor}
skim(CellChatDB$cofactor)
```

### Gene info

The final item in the database is gene annotation information.

```{r input-database-geneinfo}
skim(CellChatDB$geneInfo)
```

Network
-------

The network is a binary matrix showing interactions between proteins.

```{r input-network}
pander(as.matrix(PPI.mouse[1:5, 1:5]))
```

The full matrix has **`r nrow(PPI.mouse)`** rows and **`r ncol(PPI.mouse)`**
columns.

Create `CellChat` object
========================

**CellChat** uses a custom object which we can create from the expression matrix
and labels. We also store the database here.

```{r object}
cellchat <- createCellChat(expr)
cellchat <- addMeta(cellchat, meta = labels, meta.name = "labels")
cellchat <- setIdent(cellchat, ident.use = "labels")
cellchat@DB <- CellChatDB

cellchat
```

Pre-processing
==============

Subset signalling genes
-----------------------

First we subset the object to select only the signalling genes that are listed
in the database.

```{r pre-subset}
cellchat <- subsetData(cellchat)

cellchat
```

Identify over-expression
------------------------

The next step is to identify over-expressed genes and interactions.
Over-expressed genes are detected using a one-sided Wilcoxon test. Interactions
are selected if either a ligand or receptor is over-expressed.

```{r pre-over}
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
```

This step has identified **`r length(cellchat@var.features)`** genes and
**`r nrow(cellchat@LR$LRsig)`** interactions.

Project data
------------

The gene expression data is then projected onto a protein-protein interaction
network.

```{r pre-project}
cellchat <- projectData(cellchat, PPI.mouse)
```

Communication network inference
===============================

Compute communication probability
---------------------------------

At this stage we can now compute the communication probability and infer the
communication network. The results are stored in the `net` slot of the
`CellChat` object which contains three-dimensional arrays of probabilities and
p-values for interactions between each pair of cell types for each interaction
pair. Here are snippets of these arrays for the first interaction pair.

```{r infer-probability}
cellchat <- computeCommunProb(cellchat)

pander(cellchat@net$prob[1:5, 1:5, 1])
pander(cellchat@net$pval[1:5, 1:5, 1])
```

Infer pathway signalling
------------------------

We also compute the communication probabilities at the pathway level by
summarising all ligand-receptor interactions associated with each pathway. The
output is similar but now instead of interaction pairs we have pathways.

```{r infer-pathway}
cellchat <- computeCommunProbPathway(cellchat)

pander(cellchat@netP$prob[1:5, 1:5, 1])
```

Aggregate network
-----------------

We can get an aggregated communication network by counting the number of links
between cell types or summarising the communication probabilities. These results
are stored as two-dimensional matrices.

```{r infer-aggregate}
cellchat <- aggregateNet(cellchat)

pander(cellchat@net$count[1:5, 1:5])
pander(cellchat@net$sum[1:5, 1:5])
```

Network analysis
================

**CellChat** can perform analysis on the communication network to better
understand the roles of each cell type. Here we perform that analysis using the
pathway scores. The output is a list with a set of metrics for cell type for
each pathway.

```{r network-analysis}
cellchat <- netAnalysis_signalingRole(cellchat, slot.name = "netP")
```

The network metrics are:
`r glue_collapse(glue("**{names(cellchat@netP$centr[[1]])}**"), sep = ", ", last = " and ")`

Global patterns
===============

It may also be interesting to identify common communication patterns. We can do
this for both outgoing and incoming patterns.

```{r patterns}
cellchat <- identifyCommunicationPatterns(cellchat, pattern = "outgoing", k = 5)
cellchat <- identifyCommunicationPatterns(cellchat, pattern = "incoming", k = 5)
```

Manifold learning and classification
====================================

**CellChat** can identify groups of signalling pathways by embedding and
clustering them.

Functional similarity
----------------------

Functional similarity indicates that major senders and receivers are similar
and can be interpreted as the two pathways or interaction pairs exhibiting
similar and/or redundant roles. 

The output is a similarity matrix, UMAP embedding and cluster assignments from
$k$-means.

```{r manifold-functional}
cellchat <- computeNetSimilarity(cellchat, type = "functional", thresh = 0.25)
cellchat <- netEmbedding(cellchat, type = "functional")
cellchat <- netClustering(cellchat, type = "functional", k = 4)

pander(cellchat@netP$similarity$functional$matrix[1:5,1:5])
pander(cellchat@netP$similarity$functional$dr[1:5, ])
cellchat@netP$similarity$functional$group
```

Structural similarity
---------------------

Structural similarity is used to compare the signalling network structure,
without considering the similarity of senders and receivers.

```{r manifold-structural}
cellchat <- computeNetSimilarity(cellchat, type = "structural", thresh = 0.25)
cellchat <- netEmbedding(cellchat, type = "structural")
cellchat <- netClustering(cellchat, type = "structural", k = 4)

pander(cellchat@netP$similarity$structural$matrix[1:5,1:5])
pander(cellchat@netP$similarity$structural$dr[1:5, ])
cellchat@netP$similarity$structural$group
```

Visualisation
=============

Signalling pathways
-------------------

### Hierarchy plot

Here solid circles are sources and open circles are targets. Circle sizes are
the number of cells in each group and lines are coloured according to the source
with thicker lines indicating stronger signals.

```{r vis-pathway-hierarchy}
netVisual_aggregate(cellchat, "TGFb", vertex.receiver = 1:9, # groups to show
                    vertex.size = group_size)
```

### Circle plot

We can show a similar thing as a circle plot.

```{r vis-pathway-circle}
netVisual_aggregate(cellchat, "ncWNT", layout = "circle",
                    vertex.size = group_size)
```

Pathway contribution
--------------------

Here we show the contribution of each interaction pair to a signalling pathway.

```{r vis-contribution}
netAnalysis_contribution(cellchat, "TGFb")
```

Signalling roles
----------------

For can look at the expected role for each cell type for a pathway based on the
network analysis scores.

```{r vis-roles}
netVisual_signalingRole(cellchat, "TGFb", width = 16, height = 4,
                        font.size = 10)
```

Communication patterns
----------------------

### River plot

An alternative way to visualise the communication patterns is using a river
plot.

```{r vis-patterns-river}
netAnalysis_river(cellchat, pattern = "outgoing")
```

### Dot plot

Another alternative is a dot plot.

```{r vis-patterns-dot}
netAnalysis_dot(cellchat, pattern = "outgoing")
```

Embedding
---------

UMAP embedding of pathway similarity.

```{r umap}
netVisual_embedding(cellchat, type = "functional", pathway.remove.show = FALSE,
                    label.size = 3.5)
```

### Zoom in

```{r umap-zoom}
netVisual_embeddingZoomIn(cellchat, type = "functional", nCol = 2)
```


Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information {.unnumbered}
-------------------
