---
title: "scTHI example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the **scTHI** package and have a look at the output it produces.
More information about **scTHI** can be found at https://github.com/miccec/scTHI.

```{r libraries}
library("scTHI")
```

Input
=====

Expression
----------

The **scTHI** tutorial contains a pre-processed dataset from a glioma sample.
The sample has been filtered, transformed to log TPM and quantile normalised.

```{r input-expression}
data("H3K27", package = "scTHI.data")

pander(as.matrix(H3K27[1:5, 1:5]))
```

The matrix has **`r nrow(H3K27)`** rows (pairs) and **`r ncol(H3K27)`** columns.

Metadata
--------

We are also provided with a data frame of metadata information.

```{r input-metadata}
data("H3K27.meta", package = "scTHI.data")

skim(H3K27.meta)
```

The most important column here contains the cell type labels.

```{r input-types}
table(H3K27.meta$Type)
```

Selection
=========

For the tutorial we only use the immune cells and a subset of the malignant cells.

```{r select}
malignant <- rownames(H3K27.meta)[H3K27.meta$Type == "Malignant"][1:100]
immune    <- rownames(H3K27.meta)[H3K27.meta$Type == "Immune cell"]
```

Testing
=======

We now pass the input to the scoring function which uses an internal database of ligand-receptor pairs.
The score considers the average expression of each pair but also the percentage of cells that express each ligand/receptor.
Genes are only considered to be expressed if they are in the top 10 percent of genes in a cell.
A p-value for each interaction is computed using a permutation test.

```{r scoring}
output <- scTHI_score(
    expMat          = H3K27,
    cellCusterA     = malignant,
    cellCusterB     = immune,
    cellCusterAName = "Malignant",
    cellCusterBName = "Immune",
    topRank         = 10,
    PValue          = TRUE,
    pvalueCutoff    = 0.05,
    nPermu          = 10,
    ncore           = 1
)
```

The output is a list with **`r length(output)`** items: 
`r glue_collapse(glue("**{names(output)}**"), sep = ", ", last = " and ")`

Most of these are the inputs we provided but the scores are available in the
**result** item.

```{r results}
skim(output$result)
```

Visualisation
=============

Scores
------

We can visualise a bar plot of the scores for the significant pairs.

```{r vis-scores}
scTHI_plotResult(
    scTHIresult = output,
    cexNames    = 0.7,
    plotType    = "score"
)
```

Expression
----------

This bar plot shows the percentage of cells in each cluster that express each partner in the interaction.

```{r vis-expression}
scTHI_plotResult(
    scTHIresult = output,
    cexNames    = 0.7,
    plotType    = "pair"
)
```

t-SNE
-----

We can also look at where cells lie in a t-SNE embedding.

### Cluster

Here the cells are coloured by the selected clusters.

```{r vis-tsne-cluster}
output <- scTHI_runTsne(scTHIresult = output)

scTHI_plotCluster(
    scTHIresult = output,
    cexPoint    = 0.8,
    legendPos   = "bottomleft"
)
```

### Interactions

We can also plot the expression of specific pairs we are interested in.

```{r vis-tsne-interaction}
scTHI_plotPairs(
    scTHIresult       = output,
    cexPoint          = 0.8,
    interactionToplot = "THY1_ITGAX:ITGB2"
)
```

Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information {.unnumbered}
-------------------
