---
title: "celltalker example"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, cache = FALSE}
# Setup document
source(here::here("code", "setup.R"))

# Function dependencies
invisible(drake::readd(download_link))
```

Introduction {.unnumbered}
============

In this document we are going to run through the example analysis for the **celltalker** package and have a look at the output it produces.
More information about **celltalker** can be found at https://github.com/arc85/celltalker.

```{r libraries}
library("celltalker")
```

Input
=====

Expression
----------

The main input to **celltalker** is an expression matrix for a set of samples (3 PBMC samples and 3 tonsil samples) from GEO accession GSE139324.
We create this by reading in the raw **Cell Ranger** output files and storing them as a `Seurat` object.

> **Note:** It is not entirely clear from the **celltalker** tutorial exactly which samples to use so these results may differ from what is show there.

```{r input-expression}
data_paths <- fs::dir_ls(here("data", "celltalker"))
names(data_paths) <- c("pbmc1", "pbmc2", "pbmc3", "tonsil1", "tonsil2",
                       "tonsil3")

raw_data <- Seurat::Read10X(data_paths)
metadata <- data.frame(
    Sample = stringr::str_remove(colnames(raw_data), "_[A-Z-0-9]+")
)
metadata$Tissue <- stringr::str_remove(metadata$Sample, "[0-9]")
rownames(metadata) <- colnames(raw_data)

seurat <- Seurat::CreateSeuratObject(counts = raw_data, meta.data = metadata)

seurat
```

Ligand-receptor pairs
---------------------

**celltalker** includes a set of ligand-receptor pairs.

```{r input-lr-pairs}
skim(ramilowski_pairs)
```

**Seurat** analysis
===================

We first need to perform a standard **Seurat** analysis to get clusters and cell type labels.
Most of the details are skipped over here.

Clustering
----------

```{r seurat-cluster}
set.seed(02221989)

seurat <- Seurat::NormalizeData(seurat)
seurat <- Seurat::FindVariableFeatures(seurat)
seurat <- Seurat::ScaleData(seurat)
seurat <- Seurat::RunPCA(seurat, verbose = FALSE)
seurat <- Seurat::RunUMAP(seurat, reduction = "pca", dims = 1:15,
                          verbose = FALSE)
seurat <- Seurat::FindNeighbors(seurat, reduction = "pca", dims = 1:15,
                                verbose = FALSE)
seurat <- Seurat::FindClusters(seurat, resolution = 0.5, verbose = FALSE)

p1 <- Seurat::DimPlot(seurat, reduction = "umap", group.by = "Sample")
p2 <- Seurat::DimPlot(seurat, reduction = "umap", group.by = "Tissue")
p3 <- Seurat::DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.5",
                      label = TRUE) +
    Seurat::NoLegend()

cowplot::plot_grid(p1, p2, p3)
```

Labels
------

```{r seurat-labels}
Seurat::FeaturePlot(seurat, reduction = "umap",
                    features = c("CD3D", "CD8A", "CD4", "CD14", "MS4A1",
                                 "FCGR3A", "IL3RA"))

# Add metadata for cell types
cell_types <- vector("logical", length = ncol(seurat))
names(cell_types) <- colnames(seurat)

cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "0"] <- "CD4 T conv"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "1"] <- "CD4 T conv"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "2"] <- "B cells"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "3"] <- "B cells"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "4"] <- "CD14 monocytes"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "5"] <- "CD8"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "6"] <- "CD4 T conv"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "7"] <- "CD4 T conv"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "8"] <- "B cells"
cell_types[seurat@meta.data$RNA_snn_res.0.5 ==  "9"] <- "B cells"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "10"] <- "NK cells"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "11"] <- "CD8"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "12"] <- "Plasma"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "13"] <- "CD14 monocytes"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "14"] <- "CD16 monocytes"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "15"] <- "PDC"
cell_types[seurat@meta.data$RNA_snn_res.0.5 == "16"] <- "RBCs"

seurat[["CellType"]] <- cell_types

# Remove red blood cells
rbc_cells <- names(cell_types)[seurat@meta.data$RNA_snn_res.0.5 == "16"]
seurat <- seurat [, !colnames(seurat) %in% rbc_cells]
```

Select pairs
============

Before scoring interaction we first select a set of ligand-receptor pairs to look at.
First we check which genes are present in the dataset and then we select those were the ligand or receptor is an up-regulated marker gene for one of the tissue types.

```{r select}
ligands   <- as.character(unique(ramilowski_pairs$ligand))
receptors <- as.character(unique(ramilowski_pairs$receptor))

ligands_present   <- rownames(seurat)[rownames(seurat) %in% ligands]
receptors_present <- rownames(seurat)[rownames(seurat) %in% receptors]
lr_present <- union(ligands_present, receptors_present)

Seurat::Idents(seurat) <- "Tissue"
markers <- Seurat::FindAllMarkers(seurat, assay="RNA", features = lr_present,
                                  only.pos = TRUE, verbose = FALSE)

lr_sel <- unique(markers$gene)

interactions_sel <- rbind(
    ramilowski_pairs[ramilowski_pairs$ligand   %in% lr_sel, ],
    ramilowski_pairs[ramilowski_pairs$receptor %in% lr_sel, ]
)
```

This process selects **`r length(lr_sel)`** ligands and receptors and **`r nrow(interactions_sel)`** interaction pairs.

Rearrange data
==============

**celltalker** requires the input data in a specific structure.
Helpfully it provides some useful functions to make this easier.

Expression matrices
-------------------

The first data structure is a nested data frame with expression matrices for the selected ligands and receptors for each group, sample, and cluster.

```{r rearrange-expression}
expr_mat <- Seurat::GetAssayData(seurat, slot = "counts")
clusters <- seurat@meta.data$CellType
tissues  <- seurat@meta.data$Tissue
samples  <- seurat@meta.data$Sample

lr_matrices <- reshape_matrices(
    count.matrix          = expr_mat,
    clusters              = clusters,
    groups                = tissues,
    replicates            = samples,
    ligands.and.receptors = interactions_sel
)

# There is a big where the next step fails if there are some clusters that
# aren't present in all groups. Add dummy values to avoid this.
dummy <- pull(unnest(lr_matrices, cols = "samples"))[[1]][[1]][0, ]
lr_matrices$samples[[2]]$expr.matrices <- map(
    lr_matrices$samples[[2]]$expr.matrices, function(.exp_mats) {
        missing <- unique(clusters[!(clusters %in% names(.exp_mats))])
        for (cluster in missing) {
            .exp_mats[[cluster]] <- dummy
        }
        .exp_mats
    }
)

skim(lr_matrices)
skim(unnest(lr_matrices, cols = "samples"))
```

Expressed ligands and receptors
-------------------------------

From the nested expression matrices we then create a similar structure which contains the ligands and receptors that are expressed in each group/cluster combination.

```{r rearrange-lr}
lr_expressed <- create_lig_rec_tib(
    exp.tib               = lr_matrices,
    clusters              = clusters,
    groups                = tissues,
    replicates            = samples,
    cells.reqd            = 10,
    freq.pos.reqd         = 0.5,
    ligands.and.receptors = interactions_sel
)

skim(lr_expressed)
skim(unnest(lr_expressed[1, 2], cols = "lig.rec.exp"))
```

Interactions
============

Putative interactions
---------------------

With this data structure we can now look for putative interactions.

```{r interactions-putative}
putative_interactions <- putative_interactions(
    ligand.receptor.tibble = lr_expressed,
    clusters               = clusters,
    groups                 = tissues,
    freq.group.in.cluster  = 0.05,
    ligands.and.receptors  = interactions_sel
)

skim(putative_interactions)
```

The nested data frame returned by this function contains lists of the active ligands and receptors for each cell type in each group.

Unique interactions
-------------------

The next step identifies interactions that are unique to each group.

```{r interactions-unique}
unique_interactions <- unique_interactions(
    putative_interactions,
    group1 = "pbmc",
    group2 = "tonsil",
    interactions_sel
)

skim(unique_interactions)
```

This gives use the ligands and receptors that are unique to each group as well as those they have in common.

Visualisation
=============

Circos plot
-----------

### PBMC

```{r vis-circos-pbmc}
pbmc_lr           <- pull(unique_interactions[1, 2])[[1]]
pbmc_interactions <- pull(putative_interactions[1, 2])[[1]][pbmc_lr]

circos_plot(interactions = pbmc_interactions, clusters = clusters)
```

### Tonsil

```{r vis-circos-tonsil}
tonsil_lr           <- pull(unique_interactions[2, 2])[[1]]
tonsil_interactions <- pull(putative_interactions[2, 2])[[1]][tonsil_lr]

circos_plot(interactions = tonsil_interactions, clusters = clusters)
```

Summary {.unnumbered}
=======

Parameters {.unnumbered}
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    
)
params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files {.unnumbered}
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information {.unnumbered}
-------------------
